---
title: "Malaria_Analysis"
author: "Neema Muroki"
date: "2025-04-12"
output: pdf_document
---

```{r}
# Load the necessary packages
library(dplyr);
library(ggplot2);
library(haven);
library(tidyverse);
library(forcats);
```


```{r}
children_record<- read_dta("C:/Users/HP/Documents/R_projects/Malaria_data/KEKR81FL.DTA")
head(children_record)
```

#Explarotory Data Analysis
*It's exploration timee..*
It has a lot of columns let us extract only the most important

```{r}
# ml13e is a malaria code for malaria test 1 stands for yes child was tested for malaria 0 no the child wasn't
attr(children_record$ml13e, "labels")
```

```{r}
# Keep only relevant malaria variables
Malaria_Children <- children_record %>%
  select(caseid, scounty, szone, v024, v025, v106, v190,
         ml101, ml0, v459, v460,
         h22, h46b, h47, s406,
         ml13a:ml13z, s413a:s413c,
         hw53, hw55, hw56, hw57)

```

*getting to know my tables*
```{r}
sapply(Malaria_Children,function(x) attr(x,"label"))
```

*check different levels of the columns to better understand the data stored*

```{r}
levels(as_factor(Malaria_Children$hw57)) #anemia level
```

```{r}
levels(as_factor(Malaria_Children$hw55)) #result of measurement -hemoglobin
```
```{r}
levels(as_factor(Malaria_Children$hw53)) #"hemoglobin level (g/dl - 1 decimal)" 
```
```{r}
levels(as_factor(Malaria_Children$s406 )) #"told child had malaria"
```

```{r}
 levels(as_factor(Malaria_Children$ml101)) 
```
# Visualization

```{r}
table(as_factor(Malaria_Children$hw57)) #anemia level
```
Among children tested, 27.1% were moderately anemic, 46.2% were not anemic,24.2 were mildly anemic and 2.5% were severely anemic.

```{r}
# Create a frequency table
anemia_table <- table(as_factor(Malaria_Children$hw57))
anemia_df <- as.data.frame(anemia_table)
colnames(anemia_df) <- c("Anemia_Level", "Count")

# Plot
ggplot(anemia_df, aes(x = Anemia_Level, y = Count, fill = Anemia_Level)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("red", "orange", "yellow", "darkgreen")) +
  labs(title = "Distribution of Anemia Levels in Children",
       x = "Anemia Level",
       y = "Number of Children") +
  theme_minimal(base_size = 14)

```

# let us compare the level's of anemia level and whether they have malaria or not (told they have malaria)

```{r}
# Prepare the data
plot_data <- Malaria_Children %>%
  filter(!is.na(hw57), !is.na(s406)) %>%  # remove NAN
  mutate(
    Anemia_Level = as_factor(hw57),
    Malaria_Status = as_factor(s406)
  ) %>%
  group_by(Anemia_Level, Malaria_Status) %>%
  summarise(Count = n(), .groups = "drop")

# Plot grouped bar chart
ggplot(plot_data, aes(x = Anemia_Level, y = Count, fill = Malaria_Status)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("forestgreen", "tomato", "gray")) +
  labs(title = "Anemia Levels by Malaria Diagnosis Status",
       x = "Anemia Level",
       y = "Number of Children",
       fill = "Malaria Diagnosis") +
  theme_minimal(base_size = 14)

```
*Malaria and Anemia Are Strongly Linked*
Among children who are moderately or mildly anemic, a noticeable proportion had malaria (orange bars).

Most Non-Anemic Children Did Not Have Malaria
In the "not anemic" group, the majority had no malaria (tall green bar).
Children without anemia are less likely to have been diagnosed with malaria.

Children with anemia — especially moderate and mild anemia — are more likely to be diagnosed with malaria. Conversely, children without anemia rarely report malaria, suggesting a strong association between malaria infection and anemia severity.

# Malaria and Net 

```{r}
# Record ml101 into categories
Malaria_Children <- Malaria_Children %>%
  mutate(
    net_use = case_when(
      ml101 == 1 ~ "No Net",
      ml101 == 2 ~ "Treated Net",
      ml101 == 3 ~ "Untreated Net",
      ml101 == 4 ~ "Both Treated & Untreated",
      TRUE ~ "Unknown"
    ),
    malaria_result = as_factor(Malaria_Children$ml101)
  )

# Filter 'Unknown' 
Malaria_Children_clean <- Malaria_Children %>%
  filter(net_use != "Unknown")
```


```{r}
levels(as_factor(Malaria_Children$ml101)) 
```

```{r}
levels(as_factor(Malaria_Children$s406)) #told child had malaria 
```


```{r}
Malaria_Children <- Malaria_Children %>%
  mutate(
    malaria_result = as_factor(s406), #told child had malaria
    slept_under_net = case_when(
      ml101 == 0 ~ "No",      # 0 = no net
      ml101 > 0 ~ "Yes",      # 1, 2, 3 = some net type
      TRUE ~ NA_character_    # catch NA/don't know
    )
  )
```

# Net against Malaria diagnostics

Are children diagnosed with malaria more or less likely to have slept under a mosquito net?

```{r}
table(Malaria_Children$slept_under_net, Malaria_Children$malaria_result)
```

The proportion of malaria cases is almost the same between children who slept under nets and those who didn't.

This might seem surprising, but it could mean:

Some nets may be damaged, untreated, or improperly used

Other factors like location, mosquito resistance, or exposure times might matter more

It may indicate low statistical difference or noise in the data


```{r}
levels(as_factor(Malaria_Children$v025))#"type of place of residence"
```
v025 - place of residence
s406 - told child had malaria

```{r}
ggplot(data = Malaria_Children_clean[!is.na(Malaria_Children_clean$v025) & !is.na(Malaria_Children_clean$s406), ], aes(x = as.factor(v025), fill = as.factor(s406))) +
  geom_bar(position = "dodge") +
  labs(
    title = "Malaria Diagnosis by Place of Residence",
    x = "Type of Residence",
    y = "Diagnosis of Malaria",
    fill = "Told Child Had Malaria"
  ) +
  scale_x_discrete(labels = c("1" = "Urban", "2" = "Rural")) +
  scale_fill_manual(values = c("0" = "steelblue", "1" = "tomato"), labels = c("No", "Yes"))

```


*Most children that have malaria are from the rural areas as compared to the children in urban*

# Malaria Drugs 
*Out of all the malaria drugs taken by children with fever, what % was each drug used?*

```{r}
malaria_drugs <- Malaria_Children_clean[ c("ml13a", "ml13b", "ml13c", "ml13d", "ml13da",
                        "ml13aa", "ml13ab", "ml13e", "ml13f", "ml13g",
                        "ml13h", "ml13i", "s413a", "s413b", "s413c")]
```


```{r}
malaria_drugs <- as.data.frame(lapply(malaria_drugs, function(x) as.numeric(as.character(x))))
```

```{r}
drug_totals <- colSums(malaria_drugs == 1, na.rm = TRUE)
```


```{r}
drug_df <- data.frame(
  Drug = names(drug_totals),
  Count = as.numeric(drug_totals)
)
```


```{r}
drug_df$Drug <- recode(drug_df$Drug,
  "ml13a" = "Fansidar",
  "ml13b" = "Chloroquine",
  "ml13c" = "Amodiaquine",
  "ml13d" = "Quinine",
  "ml13da" = "Quinine Inj.",
  "ml13aa" = "Artesunate Rectal",
  "ml13ab" = "Artesunate Inj.",
  "ml13e" = "ACT Combo",
  "ml13f" = "CS Antimalarial 1",
  "ml13g" = "CS Antimalarial 2",
  "ml13h" = "CS Antimalarial 3",
  "ml13i" = "Other Antimalarial",
  "s413a" = "AL",
  "s413b" = "DHA-P",
  "s413c" = "Other ACT"
)

```


```{r}
ggplot(drug_df, aes(x = reorder(Drug, -Count), y = Count, fill = Drug)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(
    title = "Most Commonly Used Malaria Drugs",
    x = "Drug",
    y = "Number of Children Who Took the Drug"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(fill = "none")

```
AL (Artemether-Lumefantrine) was the most commonly used antimalarial drug among children with fever in the last two weeks, accounting for the highest proportion of treatments recorded in the data set.
